---
title: "Student GPA"
author:
  - name: Benjamin Loos
  - name: Andrew McRavey
  - name: Mea Johnson
date: "5/2/2026"
format:
  html:
    toc: true
execute:
  echo: false
  
editor: visual
---

# Dataset

## Raw Data
Our dataset contains data on 632 first-year students from Carnegie Mellon University, University of Washington, and Notre dame University. The relevant dataset attributes are race, gender, firstgeneration, Total sleep, midpoint sleep, Daytime Sleep, cumulative GPA, and Term GPA. 

```{r}
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv"

raw_cmu_data <- read.csv(url)
print(raw_cmu_data)

```
## tiydied Data

This is our first tidyed dataset.
```{r}
library(tidyverse)

raw <- read.csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv")

sleep_clean <- raw %>%
  select(
    id          = subject_id,
    naptime    = daytime_sleep,
    total_sleep = TotalSleepTime,
    midpoint_sleep,
    cum_gpa,
    term_gpa
  )




sleep_clean <- sleep_clean %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(midpoint_sleep),
    !is.na(naptime)
  )


glimpse(sleep_clean)
summary(sleep_clean)

```

This is out seccond tiyded dataset.
```{r}
library(tidyverse)

raw <- read_csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv")

sleep_clean2 <- raw %>%
  select(
    id          = subject_id,
    race        = demo_race,
    gender      = demo_gender,
    firstgen    = demo_firstgen,
    total_sleep = TotalSleepTime,
    university = study,
    cum_gpa,
    term_gpa,
    midpoint_sleep,
    naptime = daytime_sleep 
  )

sleep_clean2 <- sleep_clean2 %>%
  mutate(
    
    # Race: 0 = Underrepresented, 1 = Not Underrepresented
    race = factor(race,
                  levels = c(0, 1),
                  labels = c("Underrepresented", "Not Underrepresented")),
    
    # Gender: 0 = Male, 1 = Female
    gender = factor(gender,
                    levels = c(0, 1),
                    labels = c("Male", "Female")),
    
    # First-generation college student: 0 = Not First-Gen, 1 = First-Gen
    firstgen = factor(firstgen,
                      levels = c(0, 1),
                      labels = c("Not First-Gen", "First-Gen")),
   
    # Study: 1 = Carnegie Mellon University sP2018, 2 = University of Washington SP2018, 3 = University of Washington SP2019, 4 = Notre Dame University SP2016, 5 = Carnegie Mellon University SP2017
    university = factor(university,
                        levels = c(1, 2, 3, 4, 5),
                        labels = c("Carnegie Mellon SP 2018", "University of Washington SP 2018", 
                                   "University of Washington SP 2019", "Notre Dame University SP 2016", 
                                   "Carnegie Mellon University SP 2017" ))
  )

sleep_clean2 <- sleep_clean2 %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(gender),
    !is.na(firstgen),
    !is.na(race),
    !is.na(university)
  )
  
print(sleep_clean2)

```
# Questions

## what is the relationship between student GPA and their sleep?

We want to measure Student's cumulative and term GPA against their total sleep, daytime sleep, and midpoint sleep to see if there are any correlations or trends between them.

## what is the relationship between student GPA and their demographic?

Do firstgen students perform better, or vise versa? we want to know if it makes a difference or not, and what kind of difference it makes.

Does race or gender 


(idk if this is enough)

# Student GPA and Sleep

In this section, we explore the relationships between a Students GPA and several types of sleep.

## Student GPA and Their Total Sleep

### Graph

The following is a scatter plot with the y-axis as GPA and the x-axis as Total sleep. Note: Total sleep is the average sleep the student got during the semester, not the actual total amount of sleep. Each point on the graph is a single student that was randomly sampled from the data set.
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 200 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = total_sleep, y = cum_gpa)
  ) + labs(
    title = "GPA by Amount of Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "#2C6DB2") +
  geom_smooth(aes(color = "Linear Model"),method = "lm", se = FALSE)+ 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x), se = FALSE) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type")+
  theme_bw()

```

### Insights

When looking at the linear regression line, it shows a very weak fit and near zero correlation. Due to this, we can determine that there is no linear relationship/trend between a Students GPA and their total(average) sleep.

When looking at the GAM(non-linear) regression line, we can see nearly the same as the linear regression line, nearly no curve, weak fit. This tells us one of two things. That a students GPA and their average total sleep has no non-linear relation all, or the GAM Model is too simple to capture the relationship, so a more flexible model may be needed. However, due to how similar the two regression lines are, it not likely that we need a more complex model.

For this, we can conclude that simply getting more sleep will not result in a student getting a better GPA, and that getting less sleep will hurt a students GPA. There are multiple factors to that could be affecting a students GPA, like how much they study, activities outside of school, or substance abuse. There is also the possibility that quality of sleep is a more deciding factor than the total sleep.

## Student GPA and their Daytime-Sleep/Naptime

This graph is like the last one, except we replaced total sleep with the students daytime sleep, or how much they napped.

### Graph
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 200 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = naptime, y = cum_gpa)
  ) + labs(
    title = "GPA by Amount of Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "#2C6DB2") +
  geom_smooth(aes(color = "Linear Model"),method = "lm", se = FALSE)+ 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x), se = FALSE) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type")+
  theme_bw()
```

### Insights

As we can see, the linear and non-linear lines are identical. You can't even see the linear line anymore.

What does this tell us? It tells that our linear model agrees with our non-linear model, showing that it is very stable.
