---
title: "Student GPA"
author:
  - name: Benjamin Loos
  - name: Andrew McRavey
  - name: Mea Johnson
date: "5/2/2026"
format:
  html:
    toc: true
  pdf:
    toc: true
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry: 
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
  
editor: visual
---

# Data set

## Raw Data

Our data set contains data on 632 first-year students from Carnegie Mellon University, University of Washington, and Notre dame University. The relevant dataset attributes are race, gender, first-generation, Total sleep, midpoint sleep, Daytime Sleep, cumulative GPA, and Term GPA. The students' sleep data was collected via fit-bits.

```{r}
library(tidyverse)
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv"

raw_cmu_data <- read.csv(url)
glimpse(raw_cmu_data)
summary(raw_cmu_data)

```

## Tidy Data

This is our first tidy dataset.

```{r}
library(tidyverse)

raw <- read.csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv")

sleep_clean <- raw %>%
  select(
    id          = subject_id,
    naptime    = daytime_sleep,
    total_sleep = TotalSleepTime,
    midpoint_sleep,
    bedtime_mssd,
    cum_gpa,
    term_gpa
  )




sleep_clean <- sleep_clean %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(midpoint_sleep),
    !is.na(bedtime_mssd),
    !is.na(naptime)
  )


glimpse(sleep_clean)
summary(sleep_clean)

```

This is our second tidy data set.

```{r}
library(tidyverse)

raw <- read_csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv")

sleep_clean2 <- raw %>%
  select(
    id          = subject_id,
    race        = demo_race,
    gender      = demo_gender,
    firstgen    = demo_firstgen,
    total_sleep = TotalSleepTime,
    university = study,
    cum_gpa,
    term_gpa,
    midpoint_sleep,
    naptime = daytime_sleep 
  )

sleep_clean2 <- sleep_clean2 %>%
  mutate(
    
    # Race: 0 = Underrepresented, 1 = Not Underrepresented
    race = factor(race,
                  levels = c(0, 1),
                  labels = c("Underrepresented", "Not Underrepresented")),
    
    # Gender: 0 = Male, 1 = Female
    gender = factor(gender,
                    levels = c(0, 1),
                    labels = c("Male", "Female")),
    
    # First-generation college student: 0 = Not First-Gen, 1 = First-Gen
    firstgen = factor(firstgen,
                      levels = c(0, 1),
                      labels = c("Not First-Gen", "First-Gen")),
   
    # Study: 1 = Carnegie Mellon University sP2018, 2 = University of Washington SP2018, 3 = University of Washington SP2019, 4 = Notre Dame University SP2016, 5 = Carnegie Mellon University SP2017
    university = factor(university,
                        levels = c(1, 2, 3, 4, 5),
                        labels = c("Carnegie Mellon SP 2018", "University of Washington SP 2018", 
                                   "University of Washington SP 2019", "Notre Dame University SP 2016", 
                                   "Carnegie Mellon University SP 2017" ))
  )

sleep_clean2 <- sleep_clean2 %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(gender),
    !is.na(firstgen),
    !is.na(race),
    !is.na(university)
  )
  
glimpse(sleep_clean2)
summary(sleep_clean2)

```

# Questions

## what is the relationship between student GPA and their sleep?

We want to measure Student's cumulative and term GPA against their total sleep, daytime sleep, and midpoint sleep to see if there are any correlations or trends between them.

## what is the relationship between student GPA and their demographic?

Do first-gen students perform better, or vise versa? we want to know if it makes a difference or not, and what kind of difference it makes.

Does race or gender have a relationship with GPA? Is there a combination of the 2 that has a strong affect on GPA that could possibly be stronger than the sleep variables?

# Student GPA and Sleep

In this section, we explore the relationships between a Students GPA and several types of sleep.

## Student GPA and Their Total Sleep

### Graph

The following is a scatter plot with the y-axis as GPA and the x-axis as Total sleep. Note: Total sleep is the average sleep the student got during the semester, not the actual total amount of sleep. Each point on the graph is a single student that was randomly sampled from the data set.

```{r}
#| label: fig-GPA_TotalSleep
#| fig-cap: "scatterplot of the relationship between student GPA and their total sleep"
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 200 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = total_sleep, y = cum_gpa)
  ) + labs(
    title = "GPA by Amount of Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "#2C6DB2") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()

```

### Insights

When looking at the linear regression line, it shows a very weak fit and near zero correlation. Due to this, we can determine that there is no linear relationship/trend between a Students GPA and their total(average) sleep.

When looking at the GAM(non-linear) regression line, we can see nearly the same as the linear regression line, nearly no curve, weak fit. This tells us one of two things. That a students GPA and their average total sleep has no non-linear relation all, or the GAM Model is too simple to capture the relationship, so a more flexible model may be needed. However, due to how similar the two regression lines are, it is unlikely that we need a more complex model.

For this, we can conclude that there is no meaningful relationship between a students GPA and their total sleep. Simply getting more sleep will not result in a student getting a better GPA, and getting less sleep not will hurt a students GPA.

There are multiple factors to that could be affecting a students GPA, like how much they study, activities outside of school, or substance abuse. There is also the possibility that quality of sleep is a more deciding factor than the total sleep.

## Student GPA and their Daytime-Sleep/Nap-time

This graph is like the last one, except we replaced total sleep with the students daytime sleep, or how much they napped.

### Graph

```{r}
#| label: fig-GPA_DaytimeSleep
#| fig-cap: "scatterplot of the relationship between student GPA and how much they napped"
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 300 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = naptime, y = cum_gpa)
  ) + labs(
    title = "GPA by Daytime Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "yellow3") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()
```

### Insights

As we can see, the linear and non-linear lines are identical. You can't even see the linear line. Both lines show a slight negative correlation. in addition, the model is confident on the left and unsure on the right.

What does this tell us?:

The identical regression lines tells that our linear model agrees with our non-linear model, and shows the relationship is simple, no curvature, no complexity, no nonlinear pattern. In other words, we don't need a more complex model.

However, the models are very unsure for students who take more naps. Unfortunately, this is due to there being very few long nappers in the data set.There is also a very weak fit, meaning the model doesn't accurately represent the data. Even if a trend was found, it wouldn't explain the data.

With all this together, we can conclude that there is no linear, non-linear, strong, or confident relationship. Therefore, there is no meaningful relationship between a students GPA and the time they napped.

## Student GPA and Their Midpoint Sleep

Next, we graph the relationship between a students GPA and their midpoint sleep, midpoint sleep is the halfway point between when someone falls asleep and when they wake up. Midpoint sleep is relevant since it can tell us how early or late a student is sleeping. The higher value of midpoint sleep, the later the student went to bed.

### Graph

```{r}
#| label: fig-GPA_MidpointSleep
#| fig-cap: "scatterplot of the relationship between student GPA and late or early they sleep"

library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)

ggplot(sleep_clean, aes(x = midpoint_sleep, y = term_gpa)) +
  geom_point(alpha = 0.4, color = "#5A9E6F", size = 2) +
  geom_smooth(aes(color = "Linear Model"),method = "lm", se = TRUE, linewidth = 1) + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x), se = TRUE) + 
    scale_color_manual(values = c("Linear Model" = "#C44E52", "GAM Model" = "#007BFF"), name = "Regression Type") +
  labs(
    title    = "Midpoint Sleep vs Term GPA",
    subtitle = "Midpoint sleep = halfway point between bedtime and wake time(minutes after midnight)",
    x        = "Midpoint Sleep(mins)",
    y        = "Term GPA",
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray50")
  )
```

### Insights

Similar to @fig-GPA_TotalSleep and @fig-GPA_DaytimeSleep, the linear and non-linear regression lines are almost identical. We can also see that the confidence of our model is very high. The overlap and high confidence tells us that it is highly unlikely for there to be a strong non-linear relationship. However, our graph has weak fit, evidence by the scatter.

What this really means is that there is a relationship between GPA and midpoint sleep, but it is very weak and not very useful.

## Student GPA and Their Sleep Quality

Since we could not find meaningful relationships between student GPA, total sleep, daytime sleep, and midpoint sleep, we wanted to see if there we could find one the quality of sleep a student got.

Our raw data set does not have the quality of a students sleep, so we had to make a rough estimate using total sleep, daytime sleep, midpoint sleep, and bedtime_mssd(midpoint standard deviation).

$$
\text{Sleep Quality} ≈ \frac{\text{duration score} + \text{regularity score} + \text{timing score}}{3}
$$

$$
\text{duration score} = \frac{\text{nightsleep} - \text{nightmin}}{\text{nightmax} - \text{nightmin}}
$$
$$
\text{nightsleep} = \text{totalsleep} - \text{daytimesleep}
$$
$$
\text{timing score} = 1 - \frac{|\text{midpoint sleep} - \text{timing center}|}{\text{timing tol}}
$$ 

### Graph

```{r}
#| label: fig-GPA_SleepQuality
#| fig-cap: "scatterplot of the relationship between student GPA and Sleep Quality"
library(tidyverse)
library(ggplot2)

sleep_quality_vec <- function(total_sleep,
                              daytime_sleep,
                              midpoint_sleep,
                              bedtime_mssd_days,
                              night_min = 180,    # 3h -> 0
                              night_max = 540,    # 9h -> 1
                              regularity_cut_min = 0,   # 0 min -> best
                              regularity_cut_max = 300, # 300 min -> worst (0)
                              timing_center = 400,      # minutes after midnight
                              timing_tol = 240) {       # minutes for timing tolerance (0->1 mapping)
  
  # convert mssd from days to minutes
  bedtime_mssd_min <- bedtime_mssd_days * 1440
  
  # nocturnal sleep
  night_sleep <- total_sleep - daytime_sleep
  
  # duration score: linear 0->1 between night_min and night_max
  duration_score <- (night_sleep - night_min) / (night_max - night_min)
  duration_score <- pmin(1, pmax(0, duration_score))
  
  # regularity score: linear 1 -> 0 as bedtime_mssd_min goes from 0 -> regularity_cut_max
  regularity_score <- 1 - (bedtime_mssd_min - regularity_cut_min) / (regularity_cut_max - regularity_cut_min)
  regularity_score <- pmin(1, pmax(0, regularity_score))
  
  # timing score: 1 at timing_center, declines linearly to 0 at timing_tol
  timing_score <- 1 - (abs(midpoint_sleep - timing_center) / timing_tol)
  timing_score <- pmin(1, pmax(0, timing_score))
  
  # combine (simple average)
  sleep_quality <- (duration_score + regularity_score + timing_score) / 3
  
  return(sleep_quality)
}

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"
sleep_clean <- read.csv(url)

# Compute sleep_quality column
sleep_clean <- sleep_clean %>%
  mutate(
    sleep_quality = sleep_quality_vec(
      total_sleep = total_sleep,
      daytime_sleep = naptime,
      midpoint_sleep = midpoint_sleep,
      bedtime_mssd_days = bedtime_mssd,
      night_min = 180,
      night_max = 540,
      regularity_cut_max = 300,
      timing_center = median(midpoint_sleep, na.rm = TRUE),
      timing_tol = 240
    )
  )

set.seed(123)
sample_size <- 300 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = sleep_quality, y = cum_gpa)
  ) + labs(
    title = "GPA by Sleep Quality",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Sleep Quality(0-1)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "orange") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()

```

### Insights

Unfortunately, it looks like not much changed. We can see from the graph that the model has low fit, and identical regression lines, virtually no slope, much like the last graphs. The only thing that improved was our models certainty. Like the previous graphs, the high certainty, low fit, and identical regression lines means that our data contains no meaningful pattern, and the model is very sure that there is no pattern.

Surprisingly, it seems that sleep, whether its quantity or quality, doesn't have that big of an impact on a students GPA, implying there are much stronger factors other than sleep.

# Student GPA and demographics

In this section, we compare first generation students with non first generation students and represented with under-represented students by GPA. Also we, compare different combinations of the two, along with their correlations.

## Graph

This graph has multiple views showing amount of sleep by each measure, term GPA, and demographic combination (e.g. Not under-represented / First-Gen). Along with the graph there is a correlation table.

```{r}
#| label: fig-sleep-demographics 
#| fig-cap: "Sleep variables (total sleep, midpoint sleep, and naptime) plotted against term GPA, faceted  by combined race x first-generation status."

library(tidyverse)


url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_Clean2.csv"

sleep_clean2 <- read.csv(url)

# Combined Demographics Table

sleep_clean2 <- sleep_clean2 %>%
  mutate(race_firstgen = interaction(race, firstgen, sep = " / "))


# Computing Correlations
sleep_clean2 %>%
  group_by(race_firstgen) %>%
  summarise(
    corr_total = cor(total_sleep, term_gpa, use = "complete.obs"),
    corr_midpoint = cor(midpoint_sleep, term_gpa, use = "complete.obs"),
    corr_naptime = cor(naptime, term_gpa, use = "complete.obs")
  )

# One Plot per Demographic group 

sleep_clean2 %>%
  select(term_gpa, total_sleep, midpoint_sleep, naptime, race_firstgen) %>%
  pivot_longer(
    cols = c(total_sleep, midpoint_sleep, naptime),
    names_to = "sleep_variable",
    values_to = "sleep_value"
  ) %>%
  ggplot(aes(x = sleep_value, y = term_gpa)) +
  geom_point(alpha = 0.5, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_grid(race_firstgen ~ sleep_variable, scales = "free_x") +
  labs(
    title = "Sleep Variables vs Term GPA by Race × First-Gen Status",
    x = "Sleep Measure (minutes)",
    y = "Term GPA"
  ) +
  theme_minimal() +
    facet_grid(race_firstgen ~ sleep_variable, scales = "free_x", switch = "y") +
     theme(
       strip.text.y.left = element_text(angle = 0, size = 12, face = "bold"),
       strip.placement = "outside"
  )
```

## Correlation Table

```{r}
#| label: tbl-sleep-correlations
#| tbl-cap: "Correlations between sleep variables and term GPA, computed within each combined race x first-generation demographic group."

library(tidyverse)
library(gt)

# Load data
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_Clean2.csv"
sleep_clean2 <- read.csv(url)

# Create combined demographic variable
sleep_clean2 <- sleep_clean2 %>%
  mutate(race_firstgen = interaction(race, firstgen, sep = " / "))

# Compute correlations
corr_table <- sleep_clean2 %>%
  group_by(race_firstgen) %>%
  summarise(
    corr_total_sleep = cor(total_sleep, term_gpa, use = "complete.obs"),
    corr_midpoint_sleep = cor(midpoint_sleep, term_gpa, use = "complete.obs"),
    corr_naptime = cor(naptime, term_gpa, use = "complete.obs")
  ) %>%
  mutate(across(starts_with("corr_"), round, 3))  # round to 3 decimals

# Display as a formatted table
corr_table %>%
  gt() %>%
  tab_header(
    title = "Correlation of Sleep Variables with Term GPA",
    subtitle = "Grouped by Race × First-Generation Status"
  ) %>%
  cols_label(
    race_firstgen = "Race × First-Gen Group",
    corr_total_sleep = "Total Sleep",
    corr_midpoint_sleep = "Midpoint Sleep",
    corr_naptime = "Naptime"
  )

```

## Insights

Across all groups, students who slept more tended to have slightly higher term GPAs, but the effect was small. The strongest and most consistent pattern came from sleep timing: students with later midpoints (meaning they go to bed and wake up later) generally earned lower GPAs. This was especially true for Underrepresented First‑Gen students, who showed the steepest drop in GPA as midpoint sleep shifted later. The scatterplots back this up; the midpoint panels all slope downward, and the decline is most noticeable in that group.

Naptime didn’t show much of a relationship with GPA. Most groups had correlations close to zero, and the trend lines in the naptime panels were basically flat. Overall, the results suggest that *when* students sleep matters more than *how much* they sleep, and that these patterns vary a across demographic groups. Some groups, particularly Underrepresented First‑Gen students, seem more affected by later sleep timing, which could reflect differences in schedules, stress, or other outside factor.

# Appendix {.appendix}

## Code For Wrangling Data
```{r}
#| echo: true
#| eval: false


library(tidyverse)
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv"

raw_cmu_data <- read.csv(url)

raw <- read.csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv")

sleep_clean <- raw %>%
  select(
    id          = subject_id,
    naptime    = daytime_sleep,
    total_sleep = TotalSleepTime,
    midpoint_sleep,
    bedtime_mssd,
    cum_gpa,
    term_gpa
  )




sleep_clean <- sleep_clean %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(midpoint_sleep),
    !is.na(bedtime_mssd),
    !is.na(naptime)
  )





raw <- read_csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/cmu-sleep.csv")

sleep_clean2 <- raw %>%
  select(
    id          = subject_id,
    race        = demo_race,
    gender      = demo_gender,
    firstgen    = demo_firstgen,
    total_sleep = TotalSleepTime,
    university = study,
    cum_gpa,
    term_gpa,
    midpoint_sleep,
    naptime = daytime_sleep 
  )

sleep_clean2 <- sleep_clean2 %>%
  mutate(
    
    # Race: 0 = Underrepresented, 1 = Not Underrepresented
    race = factor(race,
                  levels = c(0, 1),
                  labels = c("Underrepresented", "Not Underrepresented")),
    
    # Gender: 0 = Male, 1 = Female
    gender = factor(gender,
                    levels = c(0, 1),
                    labels = c("Male", "Female")),
    
    # First-generation college student: 0 = Not First-Gen, 1 = First-Gen
    firstgen = factor(firstgen,
                      levels = c(0, 1),
                      labels = c("Not First-Gen", "First-Gen")),
   
    # Study: 1 = Carnegie Mellon University sP2018, 2 = University of Washington SP2018, 3 = University of Washington SP2019, 4 = Notre Dame University SP2016, 5 = Carnegie Mellon University SP2017
    university = factor(university,
                        levels = c(1, 2, 3, 4, 5),
                        labels = c("Carnegie Mellon SP 2018", "University of Washington SP 2018", 
                                   "University of Washington SP 2019", "Notre Dame University SP 2016", 
                                   "Carnegie Mellon University SP 2017" ))
  )

sleep_clean2 <- sleep_clean2 %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(gender),
    !is.na(firstgen),
    !is.na(race),
    !is.na(university)
  )
  

```
## Code For Total Sleep Plot

```{r}
#| echo: true
#| eval: false
#| ref.label: fig-GPA_TotalSleep

library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 200 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = total_sleep, y = cum_gpa)
  ) + labs(
    title = "GPA by Amount of Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "#2C6DB2") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()
```

## Code For Daytime Sleep Plot

``` {r}
#| echo: true
#| eval: false
#| ref.label: fig-GPA_DaytimeSleep

library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"

sleep_clean <- read.csv(url)

set.seed(123)
sample_size <- 300 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = naptime, y = cum_gpa)
  ) + labs(
    title = "GPA by Daytime Sleep",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Time slept(mins)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "yellow3") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()

```

## Code For Midpoint Sleep Plot

``` {r}
#| echo: true
#| eval: false
#| ref.label: "fig-GPA_MidpointSleep"

library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"

sleep_clean <- read.csv(url)

ggplot(sleep_clean, aes(x = midpoint_sleep, y = term_gpa)) +
  geom_point(alpha = 0.4, color = "#5A9E6F", size = 2) +
  geom_smooth(aes(color = "Linear Model"),method = "lm", se = TRUE, linewidth = 1) + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x), se = TRUE) + 
    scale_color_manual(values = c("Linear Model" = "#C44E52", "GAM Model" = "#007BFF"), name = "Regression Type") +
  labs(
    title    = "Midpoint Sleep vs Term GPA",
    subtitle = "Midpoint sleep = halfway point between bedtime and wake time(minutes after midnight)",
    x        = "Midpoint Sleep(mins)",
    y        = "Term GPA",
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    plot.caption  = element_text(color = "gray50")
  )
```

## Code for Quality Sleep Figure

``` {r}
#| echo: true
#| eval: false
#| ref.label: "fig-GPA_SleepQuality"

library(tidyverse)
library(ggplot2)

sleep_quality_vec <- function(total_sleep,
                              daytime_sleep,
                              midpoint_sleep,
                              bedtime_mssd_days,
                              night_min = 180,    # 3h -> 0
                              night_max = 540,    # 9h -> 1
                              regularity_cut_min = 0,   # 0 min -> best
                              regularity_cut_max = 300, # 300 min -> worst (0)
                              timing_center = 400,      # minutes after midnight
                              timing_tol = 240) {       # minutes for timing tolerance (0->1 mapping)
  
  # convert mssd from days to minutes
  bedtime_mssd_min <- bedtime_mssd_days * 1440
  
  # nocturnal sleep
  night_sleep <- total_sleep - daytime_sleep
  
  # duration score: linear 0->1 between night_min and night_max
  duration_score <- (night_sleep - night_min) / (night_max - night_min)
  duration_score <- pmin(1, pmax(0, duration_score))
  
  # regularity score: linear 1 -> 0 as bedtime_mssd_min goes from 0 -> regularity_cut_max
  regularity_score <- 1 - (bedtime_mssd_min - regularity_cut_min) / (regularity_cut_max - regularity_cut_min)
  regularity_score <- pmin(1, pmax(0, regularity_score))
  
  # timing score: 1 at timing_center, declines linearly to 0 at timing_tol
  timing_score <- 1 - (abs(midpoint_sleep - timing_center) / timing_tol)
  timing_score <- pmin(1, pmax(0, timing_score))
  
  # combine (simple average)
  sleep_quality <- (duration_score + regularity_score + timing_score) / 3
  
  return(sleep_quality)
}

url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_clean.csv"
sleep_clean <- read.csv(url)

# Compute sleep_quality column
sleep_clean <- sleep_clean %>%
  mutate(
    sleep_quality = sleep_quality_vec(
      total_sleep = total_sleep,
      daytime_sleep = naptime,
      midpoint_sleep = midpoint_sleep,
      bedtime_mssd_days = bedtime_mssd,
      night_min = 180,
      night_max = 540,
      regularity_cut_max = 300,
      timing_center = median(midpoint_sleep, na.rm = TRUE),
      timing_tol = 240
    )
  )

set.seed(123)
sample_size <- 300 #change this value if you want a different sample size
sleep_clean |>
  slice_sample(n = sample_size, replace = FALSE) |>
  ggplot(
    mapping = aes(x = sleep_quality, y = cum_gpa)
  ) + labs(
    title = "GPA by Sleep Quality",
    subtitle = sprintf("sample space of %s students",sample_size),
    x = "Sleep Quality(0-1)",
    y = "Student GPA"
  ) + geom_point(size = 2, color = "orange") +
  geom_smooth(aes(color = "Linear Model"),method = "lm") + 
  geom_smooth(aes(color = "GAM Model"),method = "gam", formula = y ~ s(x)) +
  scale_color_manual(values = c("Linear Model" = "red", "GAM Model" = "green"), name = "Regression Type") +
  theme_bw()
```

## Code For Demographics Plot

``` {r}
#| echo: true
#| eval: false
#| ref.label: "fig-sleep-demographics"

library(tidyverse)


url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_Clean2.csv"

sleep_clean2 <- read.csv(url)

# Combined Demographics Table

sleep_clean2 <- sleep_clean2 %>%
  mutate(race_firstgen = interaction(race, firstgen, sep = " / "))


# Computing Correlations
sleep_clean2 %>%
  group_by(race_firstgen) %>%
  summarise(
    corr_total = cor(total_sleep, term_gpa, use = "complete.obs"),
    corr_midpoint = cor(midpoint_sleep, term_gpa, use = "complete.obs"),
    corr_naptime = cor(naptime, term_gpa, use = "complete.obs")
  )

# One Plot per Demographic group 

sleep_clean2 %>%
  select(term_gpa, total_sleep, midpoint_sleep, naptime, race_firstgen) %>%
  pivot_longer(
    cols = c(total_sleep, midpoint_sleep, naptime),
    names_to = "sleep_variable",
    values_to = "sleep_value"
  ) %>%
  ggplot(aes(x = sleep_value, y = term_gpa)) +
  geom_point(alpha = 0.5, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_grid(race_firstgen ~ sleep_variable, scales = "free_x") +
  labs(
    title = "Sleep Variables vs Term GPA by Race × First-Gen Status",
    x = "Sleep Measure (minutes)",
    y = "Term GPA"
  ) +
  theme_minimal() +
    facet_grid(race_firstgen ~ sleep_variable, scales = "free_x", switch = "y") +
     theme(
       strip.text.y.left = element_text(angle = 0, size = 12, face = "bold"),
       strip.placement = "outside"
  )
```

## Code for Correlation Table

``` {r}
#| echo: true
#| eval: false
#| ref.label: "tbl-sleep-correlations"

library(tidyverse)
library(gt)

# Load data
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/Data/sleep_Clean2.csv"
sleep_clean2 <- read.csv(url)

# Create combined demographic variable
sleep_clean2 <- sleep_clean2 %>%
  mutate(race_firstgen = interaction(race, firstgen, sep = " / "))

# Compute correlations
corr_table <- sleep_clean2 %>%
  group_by(race_firstgen) %>%
  summarise(
    corr_total_sleep = cor(total_sleep, term_gpa, use = "complete.obs"),
    corr_midpoint_sleep = cor(midpoint_sleep, term_gpa, use = "complete.obs"),
    corr_naptime = cor(naptime, term_gpa, use = "complete.obs")
  ) %>%
  mutate(across(starts_with("corr_"), round, 3))  # round to 3 decimals

# Display as a formatted table
corr_table %>%
  gt() %>%
  tab_header(
    title = "Correlation of Sleep Variables with Term GPA",
    subtitle = "Grouped by Race × First-Generation Status"
  ) %>%
  cols_label(
    race_firstgen = "Race × First-Gen Group",
    corr_total_sleep = "Total Sleep",
    corr_midpoint_sleep = "Midpoint Sleep",
    corr_naptime = "Naptime"
  )

```
