---
title: "Student GPA"
author:
  - name: Benjamin Loos
  - name: Andrew McRavey
  - name: Mea Johnson
date: "5/2/2026"
format:
  html:
    toc: true
execute:
  echo: false
  
editor: visual
---

# Dataset

## Raw Data
Our dataset contains data on 632 first-year students from Carnegie Mellon University, University of Washington, and Notre dame University. The relevant dataset attributes are race, gender, firstgeneration, Total sleep, midpoint sleep, Daytime Sleep, cumulative GPA, and Term GPA. 

```{r}
url <- "https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv"

raw_cmu_data <- read.csv(url)
print(raw_cmu_data)

```
## tiydied Data

This is our first tidyed dataset.
```{r}
library(tidyverse)

raw <- read.csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv")

sleep_clean <- raw %>%
  select(
    id          = subject_id,
    naptime    = daytime_sleep,
    total_sleep = TotalSleepTime,
    midpoint_sleep,
    cum_gpa,
    term_gpa
  )




sleep_clean <- sleep_clean %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(midpoint_sleep),
    !is.na(naptime)
  )


glimpse(sleep_clean)
summary(sleep_clean)

```

This is out seccond tiyded dataset.
```{r}
library(tidyverse)

raw <- read_csv("https://raw.githubusercontent.com/Stat184-Spring2026/Sec4_CP_Ben_Andrew_Mea/refs/heads/main/cmu-sleep.csv")

sleep_clean2 <- raw %>%
  select(
    id          = subject_id,
    race        = demo_race,
    gender      = demo_gender,
    firstgen    = demo_firstgen,
    total_sleep = TotalSleepTime,
    university = study,
    cum_gpa,
    term_gpa,
    midpoint_sleep,
    naptime = daytime_sleep 
  )

sleep_clean2 <- sleep_clean2 %>%
  mutate(
    
    # Race: 0 = Underrepresented, 1 = Not Underrepresented
    race = factor(race,
                  levels = c(0, 1),
                  labels = c("Underrepresented", "Not Underrepresented")),
    
    # Gender: 0 = Male, 1 = Female
    gender = factor(gender,
                    levels = c(0, 1),
                    labels = c("Male", "Female")),
    
    # First-generation college student: 0 = Not First-Gen, 1 = First-Gen
    firstgen = factor(firstgen,
                      levels = c(0, 1),
                      labels = c("Not First-Gen", "First-Gen")),
   
    # Study: 1 = Carnegie Mellon University sP2018, 2 = University of Washington SP2018, 3 = University of Washington SP2019, 4 = Notre Dame University SP2016, 5 = Carnegie Mellon University SP2017
    university = factor(university,
                        levels = c(1, 2, 3, 4, 5),
                        labels = c("Carnegie Mellon SP 2018", "University of Washington SP 2018", 
                                   "University of Washington SP 2019", "Notre Dame University SP 2016", 
                                   "Carnegie Mellon University SP 2017" ))
  )

sleep_clean2 <- sleep_clean2 %>%
  filter(
    !is.na(total_sleep),
    !is.na(cum_gpa),
    !is.na(term_gpa),
    !is.na(gender),
    !is.na(firstgen),
    !is.na(race),
    !is.na(university)
  )
  
print(sleep_clean2)

```
# Questions

## what is the relationship between student GPA and their sleep?

We want to measure Student's cumulative and term GPA against their total sleep, daytime sleep, and midpoint sleep to see if there are any correlations or trends between them.

## what is the relationship between student GPA and whether or not they are firstgen students?

do firstgen students perform better, or vise versa? we want to know if it makes a difference or not, and what kind of difference it makes.

## what is the relationship between student GPA and their race and gender?
